### 2017-11-07 submission/manuscript.Rmd
add changes based on fig 3

***Genera Previously Associated with Carcinoma Increases Relative Risk More than Alpha Diversity:***
Both fecal and tissue samples had a significantly increased RR for carcinoma but not for adenoma [Figure 3]. The relative risk for feces was greater than either evenness or Shannon diversity [Figure 2 & 3]. The relative risk did not increase when considering the total abundance or increasing number of carcinoma associated genera [Figure 3]. The RR effect size was greater for stool (RR range = `r paste(round(min(filter(select_RR_stool, !(measure %in% c("all_four", "total_four")))[, "rr"]), digits = 2))` - `r paste(round(max(filter(select_RR_stool, !(measure %in% c("all_four", "total_four")))[, "rr"]), digits = 2))`) then for tissue (RR range = `r paste(round(min(filter(select_RR_tissue, !(measure %in% c("all_four", "total_four")))[, "rr"]), digits = 2))` - `r paste(round(max(filter(select_RR_tissue, !(measure %in% c("all_four", "total_four")))[, "rr"]), digits = 2))`). This decrease may be explained by the fact that tissue samples include matched samples.

Only two measures for adenoma were significant when investigating these CRC associated genera. The first was *Porphyromonas* for feces (RR range: `r round(filter(adn_select_RR_stool, measure == "Porphyromonas")[, "ci_lb"], digits = 2)` - `r round(filter(adn_select_RR_stool, measure == "Porphyromonas")[, "ci_ub"], digits = 2)`, P-value = `r format(filter(adn_select_RR_stool, measure == "Porphyromonas")[, "pvalue"], digits = 2, nsmall = 2)`) and the second was the total relative abundance of all four CRC associated genera on tissue (RR range: `r round(filter(adn_select_RR_tissue, measure == "total_four")[, "ci_lb"], digits = 2)` - `r round(filter(adn_select_RR_tissue, measure == "total_four")[, "ci_ub"], digits = 2)`, P-value = `r format(filter(adn_select_RR_tissue, measure == "total_four")[, "pvalue"], digits = 3, nsmall = 3)`).

@@

Our study identifies clear diversity changes both at the community level and within individual genera that are present in indivdiuals with carcinoma versus those without the disease. Although there was a step wise decrease in diversity from control to adenoma to carcinoma; this did not translate into large effect sizes for the relative risk of either of these two conditions. These clear changes were not easily recapitulated in those with adenoma. Even though CRC associated genera increase the relative risk of carcinoma they do not increase the relative risk of adenoma. This information suggests that these specific genera may not be the primary members of the microbial community that contributes to the formation of an adenoma but is for a carcinoma. Additionally, our data shows that by using the whole community our models perform better then when they only use the CRC associated genera. CRC associated genera are clearly important to carcinoma but the context or community in which these microbes are a part of can drastically increase the ability of models to make predictions. This data supports the concept that small localized changes within the community may be occuring that are important in the disease progression of colorectal cancer and that they may not directly involve CRC associated genera.

The driver-passenger model of the microbial role in CRC, as summarized by Flynn [@flynn_metabolic_2016], can be supported with this data for carcinoma but not necessarily for adenoma. The drasitically increased relative risk of disease when considering the CRC associated genera is highly supportive of this type of process. In a driver-passenger scenario it is possible that simply having the driver present or only identifying the passenger is a good enough proxy that the event is occuring. This would account for the observation that there is no synergistic increase in relative risk when accounting for either the total number or increasing abundance of these genera. The initial establishment of the driver within the system is also dependent on the community that is present and this is supported by the observation that when adding the community context to our models along with the CRC associated genera the model AUC increases.

Our carcinoma observations fit the driver-passenger model and support this concept within the framework of the transition from adenoma to carcinoma. In contrast, with the present data we can only suggest that the adenoma observations might fit with this model but the changes that occur at this timepoint are small and possibly focal to the adenoma or specific location. The stepwise decrease in diversity suggests that the adenoma community is not normal but this change is subtle. Although there may be localized changes that do depend on the driver-passenger model, supported by the *Porphyromonas* finding in adenoma stool and the relative total abundance of all four CRC associated genera in tissue [Figure 3], there may be another process involved that ultimately exacerbates the condition from a subtle localized change to a global community one. The poor performance of the Random Forest models for classifying adenoma based only on the microbiota would suggest that this is the case. It is possible to hypothesize that at early stages of the diease, how the host interacts to these subtle changes could be the catlyst that ultimately leads to this larger global dysfunctional community.

Although there are still questions that need to be answered for the microbiota and carcinoma a clearer framework is beginning to develop as to how this occurs. The role of the microbiota in adenoma is still not clear and part of the reason may be because many studies are not powered effectively to observe the small changes reported here. It is realistic to suspect that many changes in carcinoma could easily result in effect sizes that are 30% or more between the case and control. Most of the studies analyzed have sufficient power to detect these changes. In contrast, our data suggests that the adenoma effect size is relatively small. None of the studies analyzed were properly powered to detect a 10% or lower change between case and controls and this may well be the range in which differences occur in adenoma. Future studies investigating adenoma and the microbiota need to take these factors into consideration if we are to work out the role of the microbiota in adenoma genesis.

@@

By aggregating together a large collection of studies from both feces and tissue we are able to provide information in support of the driver-passenger model in the context of carcinoma. However, within the context of adenoma it is less clear that this relationship exists. These observations highlight the importance of power and sample number considerations when considering investigations into the microbiota and adenoma due to the subtle changes in community. This study helps to identify the problems that have been solved and the challenges that lie ahead in the invesitgation of colorectal cancer and the microbiota.

>>>>

***Genera Previously Associated with Carcinoma Increases Relative Risk More than Alpha Diversity:***
Both fecal and tissue samples had a significantly increased RR for carcinoma but not for adenoma [Figure 3]. The relative risk for feces was greater than either evenness or Shannon diversity [Figure 2 & 3]. The relative risk of carcinoma continuously increased as individuals tested positive for more CRC associated genera [Figure 3B & 3D]. The RR effect size was greater for stool (RR range = `r paste(round(min(filter(select_RR_stool, !(measure %in% c("all_four", "total_four")))[, "rr"]), digits = 2))` - `r paste(round(max(filter(select_RR_stool, !(measure %in% c("all_four", "total_four")))[, "rr"]), digits = 2))`) then for tissue (RR range = `r paste(round(min(filter(select_RR_tissue, !(measure %in% c("all_four", "total_four")))[, "rr"]), digits = 2))` - `r paste(round(max(filter(select_RR_tissue, !(measure %in% c("all_four", "total_four")))[, "rr"]), digits = 2))`). This decrease may be explained by the fact that tissue samples include matched samples.

Two measures in stool for adenoma were significant when investigating these CRC associated genera. The first was *Porphyromonas* (P-value = `r format(filter(adn_select_RR_stool, measure == "Porphyromonas")[, "pvalue"], digits = 2, nsmall = 2)`) and the second was being positive for three CRC associated genera (P-value = `r format(filter(adn_select_inc_4_RR_stool, measure == "three_counts")[, "pvalue"], digits = 2, nsmall = 2)`) [Figure 3A]. For tissue three measures for adenoma were significant; being positive for one CRC associated genera (P-value = `r format(filter(adn_select_inc_4_RR_tissue, measure == "one_counts")[, "pvalue"], digits = 2, nsmall = 2)`), being positive for two CRC associated genera (P-value = `r format(filter(adn_select_inc_4_RR_tissue, measure == "two_counts")[, "pvalue"], digits = 2, nsmall = 2)`), and being positive for four CRC associated genera (P-value = `r format(filter(adn_select_inc_4_RR_tissue, measure == "four_counts")[, "pvalue"], digits = 2, nsmall = 2)`) [Figure 3C].

@@

Our study identifies clear diversity changes both at the community level and within individual genera that are present in indivdiuals with carcinoma versus those without the disease. Although there was a step wise decrease in diversity from control to adenoma to carcinoma; this did not translate into large effect sizes for the relative risk of either of these two conditions. Even though CRC associated genera increases the relative risk of carcinoma it does not increase the relative risk of adenoma. This information suggests that these specific genera may not be the primary members of the microbial community that contributes to the formation of an adenoma but is for carcinoma. Additionally, our data shows that by using the whole community our models perform better then when they only use the CRC associated genera. CRC associated genera are clearly important to carcinoma but the context or community in which these microbes are a part of can drastically increase the ability of models to make predictions. This data supports the concept that small localized changes within the community may be occuring that are important in the disease progression of colorectal cancer and that they may not directly involve CRC associated genera.

The driver-passenger model of the microbial role in CRC, as summarized by Flynn [@flynn_metabolic_2016], can be supported with this data for carcinoma but not necessarily for adenoma. The drasitically increased relative risk of disease when considering the CRC associated genera is highly supportive of this type of process, especially in the context of increasing relative risk with more CRC associated genera positivity. It is also possible that in a driver-passenger scenario it is possible that simply having the driver present or only identifying the passenger is a good enough proxy that the event is occuring. This would account for the observation that there is no constant additive effect on relative risk for increasing positivity. Additionally, the initial establishment of the driver within the system is also dependent on the community that is present and this is supported by the observation that when adding the community context to our models along with the CRC associated genera the model AUC increases.

Our carcinoma observations fit the driver-passenger model and support this concept within the framework of the transition from adenoma to carcinoma. In contrast, with the present data we can only suggest that the adenoma observations might fit with this model but the changes that occur at this timepoint are small and possibly focal to the adenoma or specific location. The stepwise decrease in diversity suggests that the adenoma community is not normal but this change is subtle. Although there may be localized changes that do depend on the driver-passenger model, supported by the one, two, and four positive CRC associated genera in tissue relative risk increases [Figure 3C], there may be another process involved that ultimately exacerbates the condition from a subtle localized change to a global community one. The poor performance of the Random Forest models for classifying adenoma based only on the microbiota would suggest that this is the case. It is possible to hypothesize that at early stages of the diease, how the host interacts to these subtle changes could be the catlyst that ultimately leads to this larger global dysfunctional community.

Although there are still questions that need to be answered for the microbiota and carcinoma, a clearer framework is beginning to develop as to how this occurs. The role of the microbiota in adenoma is still not clear and part of the reason this may be is because many studies are not powered effectively to observe the small changes reported here. It is realistic to suspect that many changes in carcinoma could easily result in effect sizes that are 30% or more between the case and control. Most of the studies analyzed have sufficient power to detect these type of changes. In contrast, our data suggests that the adenoma effect size is relatively small. None of the studies analyzed were properly powered to detect a 10% or lower change between case and controls and this may well be the range in which differences consistently occur in adenoma. Future studies investigating adenoma and the microbiota need to take these factors into consideration if we are to work out the role of the microbiota in adenoma formation.

@@

By aggregating together a large collection of studies from both feces and tissue we are able to provide information in support of the driver-passenger model in the context of carcinoma. However, within the context of adenoma it is less clear that this relationship exists. These observations highlight the importance of power and sample number considerations when considering investigations into the microbiota and adenoma due to the subtle changes in the community.

### 2017-11-07 submission/manuscript.tex
add changes based on fig 3

\textbf{\emph{Fecal Diversity is Lower in Those with Carcinoma and
Increases Relative Risk:}} Using power transformed and Z-score
normalized alpha diversity metrics both evenness and the Shannon
diversity metrics in feces are lower in those with carcinoma then in
controls but not for tissue samples {[}Figure 1{]}. Using linear
mixed-effects to control for study and variable region there was a
significant decrease from control to adenoma to carcinoma for both
evenness (P-value = 0.025) and Shannon diversity (P-value = 0.043). This
effect was not observed in tissue when additionally controlling for
whether the sample came from the same individual (P-value \textgreater{}
0.05). For fecal samples a decrease in Shannon diversity and evenness
resulted in a significantly increased relative risk for carcinoma
(P-value = 0.01 and P-value = 0.0011, respectively) {[}Figure 2{]}.
Although these values were significant the effect size was relatively
small for both metrics (Shannon RR = 1.31 and evenness RR = 1.34)
{[}Figure 2{]}. There was no increased relative risk for these metrics
for adenoma or for tissue in general {[}Figure S1-3{]}.

@@

\textbf{\emph{Genera Previously Associated with Carcinoma Increases
Relative Risk More than Alpha Diversity:}} Both fecal and tissue samples
had a significantly increased RR for carcinoma but not for adenoma
{[}Figure 3{]}. The relative risk for feces was greater than either
evenness or Shannon diversity {[}Figure 2 \& 3{]}. The relative risk did
not increase when considering the total abundance or increasing number
of carcinoma associated genera {[}Figure 3{]}. The RR effect size was
greater for stool (RR range = 1.78 - 2.64) then for tissue (RR range =
1.33 - 1.53). This decrease may be explained by the fact that tissue
samples include matched samples.

Only two measures for adenoma were significant when investigating these
CRC associated genera. The first was \emph{Porphyromonas} for feces (RR
range: 1.06 - 2.29, P-value = 0.023) and the second was the total
relative abundance of all four CRC associated genera on tissue (RR
range: 1.21 - 3.89, P-value = 0.00933).

@@

Our study identifies clear diversity changes both at the community level
and within individual genera that are present in indivdiuals with
carcinoma versus those without the disease. Although there was a step
wise decrease in diversity from control to adenoma to carcinoma; this
did not translate into large effect sizes for the relative risk of
either of these two conditions. These clear changes were not easily
recapitulated in those with adenoma. Even though CRC associated genera
increase the relative risk of carcinoma they do not increase the
relative risk of adenoma. This information suggests that these specific
genera may not be the primary members of the microbial community that
contributes to the formation of an adenoma but is for a carcinoma.
Additionally, our data shows that by using the whole community our
models perform better then when they only use the CRC associated genera.
CRC associated genera are clearly important to carcinoma but the context
or community in which these microbes are a part of can drastically
increase the ability of models to make predictions. This data supports
the concept that small localized changes within the community may be
occuring that are important in the disease progression of colorectal
cancer and that they may not directly involve CRC associated genera.

The driver-passenger model of the microbial role in CRC, as summarized
by Flynn {[}2{]}, can be supported with this data for carcinoma but not
necessarily for adenoma. The drasitically increased relative risk of
disease when considering the CRC associated genera is highly supportive
of this type of process. In a driver-passenger scenario it is possible
that simply having the driver present or only identifying the passenger
is a good enough proxy that the event is occuring. This would account
for the observation that there is no synergistic increase in relative
risk when accounting for either the total number or increasing abundance
of these genera. The initial establishment of the driver within the
system is also dependent on the community that is present and this is
supported by the observation that when adding the community context to
our models along with the CRC associated genera the model AUC increases.

Our carcinoma observations fit the driver-passenger model and support
this concept within the framework of the transition from adenoma to
carcinoma. In contrast, with the present data we can only suggest that
the adenoma observations might fit with this model but the changes that
occur at this timepoint are small and possibly focal to the adenoma or
specific location. The stepwise decrease in diversity suggests that the
adenoma community is not normal but this change is subtle. Although
there may be localized changes that do depend on the driver-passenger
model, supported by the \emph{Porphyromonas} finding in adenoma stool
and the relative total abundance of all four CRC associated genera in
tissue {[}Figure 3{]}, there may be another process involved that
ultimately exacerbates the condition from a subtle localized change to a
global community one. The poor performance of the Random Forest models
for classifying adenoma based only on the microbiota would suggest that
this is the case. It is possible to hypothesize that at early stages of
the diease, how the host interacts to these subtle changes could be the
catlyst that ultimately leads to this larger global dysfunctional
community.

Although there are still questions that need to be answered for the
microbiota and carcinoma a clearer framework is beginning to develop as
to how this occurs. The role of the microbiota in adenoma is still not
clear and part of the reason may be because many studies are not powered
effectively to observe the small changes reported here. It is realistic
to suspect that many changes in carcinoma could easily result in effect
sizes that are 30\% or more between the case and control. Most of the
studies analyzed have sufficient power to detect these changes. In
contrast, our data suggests that the adenoma effect size is relatively
small. None of the studies analyzed were properly powered to detect a
10\% or lower change between case and controls and this may well be the
range in which differences occur in adenoma. Future studies
investigating adenoma and the microbiota need to take these factors into
consideration if we are to work out the role of the microbiota in
adenoma genesis.

@@

\textbf{Figure 1: }

\textbf{Figure 2: }

\textbf{Figure 3: }

@@

\textbf{Figure S1: }

>>>>

\textbf{\emph{Fecal Diversity is Lower in Those with Carcinoma and
Increases Relative Risk:}} Using power transformed and Z-score
normalized alpha diversity metrics both evenness and the Shannon
diversity metrics in feces are lower in those with carcinoma then in
controls but not for tissue samples {[}Figure 1{]}. Using linear
mixed-effects to control for study and variable region there was a
significant decrease from control to adenoma to carcinoma for both
evenness (P-value = 0.025) and Shannon diversity (P-value = 0.043). This
effect was not observed in tissue when additionally controlling for
whether the sample came from the same individual (P-value \textgreater{}
0.05). For fecal samples a decrease in Shannon diversity and evenness
resulted in a significantly increased relative risk for carcinoma
(P-value = 0.01 and P-value = 0.0011, respectively) {[}Figure 2{]}.
Although these values were significant the effect size was relatively
small for both metrics (Shannon RR = 1.31 and evenness RR = 1.34)
{[}Figure 2{]}. There was no increased relative risk for these metrics
for adenoma or for tissue in general {[}Figure 2A \& S1{]}.

@@

\textbf{\emph{Genera Previously Associated with Carcinoma Increases
Relative Risk More than Alpha Diversity:}} Both fecal and tissue samples
had a significantly increased RR for carcinoma but not for adenoma
{[}Figure 3{]}. The relative risk for feces was greater than either
evenness or Shannon diversity {[}Figure 2 \& 3{]}. The relative risk of
carcinoma continuously increased as individuals tested positive for more
CRC associated genera {[}Figure 3B \& 3D{]}. The RR effect size was
greater for stool (RR range = 1.61 - 2.74) then for tissue (RR range =
1.21 - 1.81). This decrease may be explained by the fact that tissue
samples include matched samples.

Two measures in stool for adenoma were significant when investigating
these CRC associated genera. The first was \emph{Porphyromonas} (P-value
= 0.023) and the second was being positive for three CRC associated
genera (P-value = 0.022) {[}Figure 3A{]}. For tissue three measures for
adenoma were significant; being positive for one CRC associated genera
(P-value = 0.032), being positive for two CRC associated genera (P-value
= 0.008), and being positive for four CRC associated genera (P-value =
0.039) {[}Figure 3C{]}.

@@

Our study identifies clear diversity changes both at the community level
and within individual genera that are present in indivdiuals with
carcinoma versus those without the disease. Although there was a step
wise decrease in diversity from control to adenoma to carcinoma; this
did not translate into large effect sizes for the relative risk of
either of these two conditions. Even though CRC associated genera
increases the relative risk of carcinoma it does not increase the
relative risk of adenoma. This information suggests that these specific
genera may not be the primary members of the microbial community that
contributes to the formation of an adenoma but is for carcinoma.
Additionally, our data shows that by using the whole community our
models perform better then when they only use the CRC associated genera.
CRC associated genera are clearly important to carcinoma but the context
or community in which these microbes are a part of can drastically
increase the ability of models to make predictions. This data supports
the concept that small localized changes within the community may be
occuring that are important in the disease progression of colorectal
cancer and that they may not directly involve CRC associated genera.

The driver-passenger model of the microbial role in CRC, as summarized
by Flynn {[}2{]}, can be supported with this data for carcinoma but not
necessarily for adenoma. The drasitically increased relative risk of
disease when considering the CRC associated genera is highly supportive
of this type of process, especially in the context of increasing
relative risk with more CRC associated genera positivity. It is also
possible that in a driver-passenger scenario it is possible that simply
having the driver present or only identifying the passenger is a good
enough proxy that the event is occuring. This would account for the
observation that there is no constant additive effect on relative risk
for increasing positivity. Additionally, the initial establishment of
the driver within the system is also dependent on the community that is
present and this is supported by the observation that when adding the
community context to our models along with the CRC associated genera the
model AUC increases.

Our carcinoma observations fit the driver-passenger model and support
this concept within the framework of the transition from adenoma to
carcinoma. In contrast, with the present data we can only suggest that
the adenoma observations might fit with this model but the changes that
occur at this timepoint are small and possibly focal to the adenoma or
specific location. The stepwise decrease in diversity suggests that the
adenoma community is not normal but this change is subtle. Although
there may be localized changes that do depend on the driver-passenger
model, supported by the one, two, and four positive CRC associated
genera in tissue relative risk increases {[}Figure 3C{]}, there may be
another process involved that ultimately exacerbates the condition from
a subtle localized change to a global community one. The poor
performance of the Random Forest models for classifying adenoma based
only on the microbiota would suggest that this is the case. It is
possible to hypothesize that at early stages of the diease, how the host
interacts to these subtle changes could be the catlyst that ultimately
leads to this larger global dysfunctional community.

Although there are still questions that need to be answered for the
microbiota and carcinoma, a clearer framework is beginning to develop as
to how this occurs. The role of the microbiota in adenoma is still not
clear and part of the reason this may be is because many studies are not
powered effectively to observe the small changes reported here. It is
realistic to suspect that many changes in carcinoma could easily result
in effect sizes that are 30\% or more between the case and control. Most
of the studies analyzed have sufficient power to detect these type of
changes. In contrast, our data suggests that the adenoma effect size is
relatively small. None of the studies analyzed were properly powered to
detect a 10\% or lower change between case and controls and this may
well be the range in which differences consistently occur in adenoma.
Future studies investigating adenoma and the microbiota need to take
these factors into consideration if we are to work out the role of the
microbiota in adenoma formation.

@@

\textbf{Figure 1: Alpha Diversity Differences between Control, Adenoma,
and Carcinoma Across Sampling Site.} A) Alpha diversity metric
differences by group in stool samples. B) Alpha diversity metric
differences by group in unmatched tissue samples. C) Alpha diversity
metric differences by group in matched tissue samples. The dashed line
represents a Z-score of 0 or no difference from the median.

\textbf{Figure 2: Relative Risk for Adenoma or Carcinoma based on Alpha
Diversity Metrics in Stool.} A) Alpha metric relative risk for adenoma.
B) Alpha metric relative risk for carcinoma. Colors represent the
different variable regions used within the respective study.

\textbf{Figure 3: Colorectal Cancer Associated Genera Relative Risk for
Adenoma and Carcinoma in Stool and Tissue.} A) Adenoma relative risk in
stool. B) Carinoma relative risk in stool. C) Adenoma relative risk in
tissue. D) Carcinoma relative risk in tissue. For all panels the
relative risk was also compared to whether one, two, three, or four of
the CRC associated genera were present.

@@

\textbf{Figure S1: Relative Risk for Adenoma or Carcinoma based on Alpha
Diversity Metrics in Tissue.} A) Alpha metric relative risk for adenoma.
B) Alpha metric relative risk for carcinoma. Colors represent the
different variable regions used within the respective study.
